---
title: "TidyVerse - Extension to Aaron Zalki's assignment"
author: "Don Padmapepruma"
date: "12/1/2019"
output:
  html_document:
    toc: True
    toc_float: true
---

## Raw Drug-Use Data
### Background on the Data

I am tidying a fivethirtyeight dataset on drug use [Github](https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age). I am transforming the columns of drugs (3 to 28) into rows, in order to perform analysis on the different drugs used by different age groups.
It is important to note that any column header in the dataset with the word 'use' is referring to a percentage of those in an age group who have used the specific drug in the past 12 months. 
Any column header in the dataset with the word 'frequency' is referring to the median number of times a user in an age group used the specific drug in the past 12 months.
For this analysis, I am only interested in the percentage the drug is used among different age groups.


### Loading Data

```{r raw data, include=TRUE}
library(RCurl)
raw_drug <- read.csv(text=getURL("https://raw.githubusercontent.com/fivethirtyeight/data/master/drug-use-by-age/drug-use-by-age.csv"), header=TRUE, stringsAsFactors = FALSE)
head (raw_drug)

```

## Tidy & Transform


I am using the packages tidyr and dplyr to change the raw data to its final form. 

```{r transforming, include=TRUE}
library(tidyr)
library(dplyr)
drug_gather <- gather(raw_drug, "Drug", "Percent", 3:28)
drug_gather$Percent <- as.numeric(drug_gather$Percent)
drug_separate <- separate(drug_gather, "Drug", c("Drug", "Usage"))
drug_filter <- filter(drug_separate, Usage=='use')

names (drug_filter) <- c("Age", "Size", "Drug", "Usage", "Percent")

head(drug_filter)
```
## Analysis

### Teenagers

I am using the package ggplot to illustrate the differences found in drug use between teenagers and young adults.

```{r teen, include=TRUE}
library(ggplot2)
teen <- ggplot(subset(drug_filter, Age %in% c("13","14","15", "16", "17", "18","19")), 
aes(x=Drug, y=Percent, fill=Age)) + geom_bar(stat="identity", position=position_dodge()) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

teen

```

### Young Adults

```{r young adult, include=TRUE}
young_adults <- ggplot(subset(drug_filter, Age %in% c("18","19","20","21","22-23","24-25")), 
aes(x=Drug, y=Percent, fill=Age)) + geom_bar(stat="identity", position=position_dodge()) + theme(axis.text.x = element_text(angle = 90, hjust = 1))


young_adults
```

## Task 2 by Don P.

### Introduction  

I will be working on the original dataset for my part of the extention. The data set **Drug Use by Age** is part of the article **[How Baby Boomers Get High](https://fivethirtyeight.com/features/how-baby-boomers-get-high/)** It covers 13 drugs across 17 age groups. 
Source of the data: [National Survey on Drug Use and Health from the Substance Abuse and Mental Health Data Archive](https://www.icpsr.umich.edu/icpsrweb/content/SAMHDA/index.html)



```{r}
raw_drug
```

### tidying the datafram

I decided to rename some columns using rename() function.   

```{r}
drug_usage <- raw_drug %>%
  rename(sample.size = n, alcohol = alcohol.use, marijuana = marijuana.use, cocaine = cocaine.use,crack = crack.use, heroin = heroin.use, hallucinogen = hallucinogen.use, inhalant = inhalant.use, pain.releiver = pain.releiver.use, oxycontin = oxycontin.use, tranquilizer=tranquilizer.use, stimulant=stimulant.use, meth=meth.use, sedative=sedative.use)
head(drug_usage)
```

As my second step I got rid of drug frequency columns using select() function as I only wanted to keep the drug usage column for my further analysis.  

```{r}
library(dplyr)
drug_usage <- drug_usage %>%
  select(-c(4, 6,8,10,12,14,16,18, 20, 22,24,26,28))
head(drug_usage)
```

For my next step I used gather() function to make my wide data into key-value pairs of **Drug** and **Percent**    

```{r}
tidy_drug_usage <- drug_usage %>%
  gather("Drug", "Percent", c(3:15))

head(tidy_drug_usage)
```

### Visualization  

For the easy visualization of the data that I just tidied, I am using age variable in three categories.  
1. Teens: 13-19  
2. Young Adults: 20-25  
3. Adults: 26-65+  

```{r}
Teens <- ggplot(subset(drug_filter, Age %in% c("13","14","15","16","17","18", "19")), 
aes(x=Drug, y=Percent, fill=Age)) + geom_bar(stat="identity", position=position_dodge()) + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
coord_flip()

Teens
```

```{r}
library(ggplot2)
young_adults <- ggplot(subset(drug_filter, Age %in% c("20","21","22-23","24-25")), 
aes(x=Drug, y=Percent, fill=Age)) + geom_bar(stat="identity", position=position_dodge()) + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
coord_flip()

young_adults
```


```{r}
Adults <- ggplot(subset(drug_filter, Age %in% c("26-29","30-34","35-49","50-64","65+")), 
aes(x=Drug, y=Percent, fill=Age)) + geom_bar(stat="identity", position=position_dodge()) + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
coord_flip()

Adults
```

conclution: Just by looking at these plots we can conclude that for all three age categories, alcohol is the highest consumed drug. And Marijuana makes the second highest in the list for all three age categories.
